### 2020-07-20 misc/expo/genfcts.tex
Added Anna to homies

\def\Seq{\op{\tensc Seq}}
\def\Set{\op{\tensc Set}}
\def\Cyc{\op{\tensc Cyc}}
\def\A{{\cal A}}
\def\B{{\cal B}}
\def\C{{\cal C}}
\def\D{{\cal D}}
\def\E{{\cal E}}
\def\F{{\cal F}}
\def\I{{\cal I}}
\def\M{{\cal M}}
\def\P{{\cal P}}
\def\T{{\cal T}}
\def\Z{{\cal Z}}
\def\bul{\bullet}

@@

\maketitle{A crash course on generating functions}{by}{Marcel K. Goh}{10 July 2020}

\section\the\sectcount. Introduction

When counting discrete structures, we often run into sequences that are recursively defined.
For example, suppose we have the recurrence
$$R_n = \cases{1,&if $n=0$; \cr 2R_{n-1}+3, & if $n\geq 1$.}$$
We might want to know how the sequence $(R_n)$ grows as $n$ gets large.
Now, this particular recurrence
is easy enough to solve by the substitution method. But often, it is easier to work with the
{\it generating function} of a sequence. For a sequence $(A_n)$, this is the formal power series
$$A(z) = \sum_{n\geq 0} A_nz^n.$$
In his textbook, titled {\sl generatingfunctionology},
H. S. Wilf famously wrote: ``A generating function is a clothesline on which we hang up a sequence
of numbers for display.'' That is exactly how we should treat the function $A(z)$ for now;
in later sections, we will discuss how this series can be treated as an analytic object.

@@

\boldlabel Fibonacci numbers. Let $\I^*$ denote the cute combinatorial class $\{1,2\}$,
which clearly has the generating function $I^*(z) = z + z^2$. Let $\T$ be the set of partitions
using only the numbers 1 and 2 (for simplicity's sake, we also include the neutral object, i.e.\ $T_0=1$).
We have
$$\eqalign{
1 &= 1, \cr
2 &= 1+1 = 2,\cr
3 &= 1 + 1 + 1 = 1 + 2 = 2+1, \cr
4 &= 1 + 1 + 1 + 1 = 1 + 1 + 2 = 1 + 2 + 1 = 2 + 1 + 1 = 2 + 2,\cr
}$$
and so on. After computing a few more terms, we notice the pattern
$$(T_n)_{n\geq 0} = (1,1,2,3,5,8,13,\ldots).$$
This is almost the infamous Fibonacci sequence; we have $T_n = F_{n+1}$.
Hence $T(z)$ must be the generating function
of the Fibonacci numbers, divided by $z$.
We can easily derive this function with the symbolic method, since a composition
using only 1s and 2s is simply a sequence of elements of $\I^*$. We conclude that
$$\T = \Seq(\I^*)\qquad\hbox{and}\qquad T(z) = {1\over 1-I^*(z)} = {1\over 1-z-z^2}.\adveq,$$
so we know that the generating function of the Fibonacci numbers is
\newcount\fibeq
\fibeq=\eqcount
$$F(z) = {z\over 1-z-z^2}.\adveq$$
\medskip

@@

\boldlabel Cayley trees. We are now equipped to count rooted non-plane labelled trees, sometimes called
Cayley trees. Children of a given node are not ordered and nodes in these trees have unrestricted degree,
so we have the specification
$$\T = \Z \star \Set(\T).$$
This immediately gives the functional equation
$$T(z) = ze^{T(z)}$$
and by Lagrange's inversion theorem, with $f(u) = u/e^u$ and $H(u) = u$, we obtain
$$n![z^n]T(z) = n!\bigg({1\over n}[u^{n-1}]e^{un}\bigg)
              = n!\cdot{1\over n}\cdot{n^{n-1}\over(n-1)!} = n^{n-1}.$$
Note that there are $n^{n-2}$ {\it unrooted} Cayley trees, since there are $n$ choices for the root.
This is called Cayley's formula and the standard combinatorial proof involves establishing
a bijection with Pr\"ufer sequences; see Pr\"ufer (1918).
\medskip

@@

\boldlabel Cycles in a permutation. Recall that a permutation is a set of cycles. If we mark every cycle
with a $u$, we get the specification
$$\P = \Set(u\Cyc(\Z)),$$
which gives the bivariate exponential generating function
$$P(z,u) = \exp\bigg(u\log{1\over 1-z}\bigg) = \exp\bigg(\log{1\over (1-z)^u}\bigg) = \bigg({1\over 1-z}\bigg)^u\adveq$$
for the number of $n$-letter permutations with exactly $k$ cycles. We can calculate the partial derivatives
$$P_u(z,u) = \bigg({1\over 1-z}\bigg)^u\log{1\over 1-z}$$
Here the random variable $\chi$ counts the number of cycles, and we have
$$\ex_{\P_n}\{\chi\} = {[z^n] P_u(z,u)\eval_{u=1} \over [z^n]P(z,1)} = 1 + {1\over 2} + \cdots + {1\over n} = H_n.$$
The harmonic numbers $H_n$ can be expanded as $\log n + \gamma + O(1/n)$, so we find that there are roughly
$\log n$ cycles in a random permutation on $n$ letters.
A slightly harder analysis of Taylor coefficients retrives the second moment.
From this, it can be shown that $\var_{\P_n}\{\chi\}\sim \log n$ an well.
\medskip

\boldlabel External nodes in a binary tree. Hidden inside the specification for binary trees,
we can see that we are actually only counting internal nodes, since we have given $\E$, the external node,
the generating function of 1 (meaning it has size 0). If we mark it with a $u$,
we arrive at
$$\B = u\E + \Z\times\B\times\B$$
and the bivariate ordinary generating function satisfies
$$B(z,u) = u + zB(z)^2.$$
Using the quadratic formula once again, we find that the generating function is
$$B(z,u) = {1-\sqrt{1-4uz} \over 2z},$$
which checks out, since setting $u=1$ gives exactly \refeq{\the\catalaneq}. We have the partial derivatives
$$P_u(z,u) = {1\over \sqrt{1-4uz}} \qquad\hbox{and}\qquad P_u^2(z,u) = {2z \over (1-4uz)^{3/2}}.$$
With the substitution $u=1$,
these are both relatively well-known generating functions. The first is the generating function
for the central binomial coefficients ${2n\choose n}$ and
the second is $z {\partial\over \partial z} P_u(z,1) $, so it has coefficients $n{2n\choose n}$
(the powers of $n$ drop down into the coefficient when differentiating, and multiplying by $z$ restores
the correct power).
We calculated in \refeq{\the\catalanformula} that the
total number of binary trees with $n$ internal nodes is ${2n\choose n}/(n+1)$, so applying our method gives us
$$\ex_{\B_n}\{\chi\} = {{2n\choose n}\over {1\over n+1}{2n\choose n}} = n+1$$
and
$$\ex_{\B_n}\{\chi^2\} = {n{2n\choose n} \over {1\over n+1}{2n\choose n}} + n + 1 = n(n+1)+n+1 = (n+1)^2.$$
So the average number of external nodes in a tree with $n$ internal nodes is $n+1$ and since
$$\var_{\B_n}\{\chi\} = \ex_{\B_n}\{\chi^2\} - \ex_{\B_n}\{\chi\}^2 = (n+1)^2 - (n+1)^2 = 0,$$
we discover that, in fact, every tree with $n$ internal nodes has exactly $n+1$ external nodes.
(We have performed what is probably the most convoluted proof of this simple fact, but it is a good exercise
in calculating moments using bivariate generating functions.)

\section References
\frenchspacing

>>>>

\def\Seq{\op{\tensc Seq}}
\def\Set{\op{\tensc Set}}
\def\Cyc{\op{\tensc Cyc}}
\def\Res{\op{\rm Res}}
\def\A{{\cal A}}
\def\B{{\cal B}}
\def\C{{\cal C}}
\def\D{{\cal D}}
\def\E{{\cal E}}
\def\F{{\cal F}}
\def\I{{\cal I}}
\def\M{{\cal M}}
\def\P{{\cal P}}
\def\T{{\cal T}}
\def\Z{{\cal Z}}
\def\bul{\bullet}

\def\d{\,d}

\def\CC{{\bf C}}
\def\RR{{\bf R}}
\def\eps{\epsilon}

@@

\maketitle{A crash course on generating functions}{by}{Marcel K. Goh}{17 July 2020}

\bigskip
{\obeylines\eightssi
\hfill A generating function is a clothesline
\hfill on which we hang up a sequence for display.
\eightss
\smallskip
\hfill --- HERBERT S. WILF, {\eightssi generatingfunctionology} (1989)}
\section\the\sectcount. Introduction

When counting discrete structures, we often run into sequences that are recursively defined.
For example, suppose we have the recurrence
$$R_n = \cases{1,&if $n=0$; \cr 2R_{n-1}+3, & if $n\geq 1$.}$$
We might want to know how the sequence $(R_n)$ grows as $n$ gets large.
Now, this particular recurrence
is easy enough to solve by the substitution method. But often, it is easier to work with the
{\it generating function} of a sequence. For a sequence $(A_n)$, this is the formal power series
$$A(z) = \sum_{n\geq 0} A_nz^n.$$
H. S. Wilf famously described a generating function as a clothesline, and
that is exactly how we should treat the function $A(z)$ for now;
in later sections, we will discuss how this series can be treated as an analytic object.

@@

\boldlabel Fibonacci numbers. Let $\I^*$ denote the cute combinatorial class $\{1,2\}$,
which clearly has the generating function $I^*(z) = z + z^2$. Let $\T$ be the set of partitions
using only the numbers 1 and 2 (for simplicity's sake, we also include the neutral object, i.e.\ $T_0=1$).
We have
$$\eqalign{
1 &= 1, \cr
2 &= 1+1 = 2,\cr
3 &= 1 + 1 + 1 = 1 + 2 = 2+1, \cr
4 &= 1 + 1 + 1 + 1 = 1 + 1 + 2 = 1 + 2 + 1 = 2 + 1 + 1 = 2 + 2,\cr
}$$
and so on. After computing a few more terms, we notice the pattern
$$(T_n)_{n\geq 0} = (1,1,2,3,5,8,13,\ldots).$$
This is almost the infamous Fibonacci sequence; we have $T_n = F_{n+1}$.
Hence $T(z)$ must be the generating function
of the Fibonacci numbers, divided by $z$.
We can easily derive this function with the symbolic method, since a composition
using only 1s and 2s is simply a sequence of elements of $\I^*$. We conclude that
$$\T = \Seq(\I^*)\qquad\hbox{and}\qquad T(z) = {1\over 1-I^*(z)} = {1\over 1-z-z^2},\adveq$$
and the generating function of the Fibonacci numbers is
\newcount\fibeq
\fibeq=\eqcount
$$F(z) = {z\over 1-z-z^2}.\adveq$$
\bigskip
{\obeylines\eightssi
\hfill ``What's one and one and one and one and one
\hfill and one and one and one and one and one?''
\hfill ``I don't know,'' said Alice. ``I lost count.''
\hfill ``She can't do Addition,'' the Red Queen interrupted.
\eightss
\smallskip
\hfill --- LEWIS CARROLL, {\eightssi Through the Looking Glass} (1871)}
\bigskip

@@

\boldlabel Cayley trees. We are now equipped to count rooted non-plane labelled trees, sometimes called
Cayley trees. Children of a given node are not ordered and nodes in these trees have unrestricted degree,
so we have the specification
$$\T = \Z \star \Set(\T).$$
This immediately gives the functional equation
$$T(z) = ze^{T(z)}$$
and by Lagrange's inversion theorem, with $f(u) = u/e^u$ and $H(u) = u$, we obtain
\newcount\cayleyexact
\cayleyexact=\eqcount
$$n![z^n]T(z) = n!\bigg({1\over n}[u^{n-1}]e^{un}\bigg)
              = n!\cdot{1\over n}\cdot{n^{n-1}\over(n-1)!} = n^{n-1}.\adveq$$
Note that there are $n^{n-2}$ {\it unrooted} Cayley trees, since there are $n$ choices for the root.
This is called Cayley's formula and the standard combinatorial proof involves establishing
a bijection with Pr\"ufer sequences; see Pr\"ufer (1918).
\medskip

@@

\boldlabel Cycles in a permutation. Recall that a permutation is a set of cycles. If we mark every cycle
with a $u$, we get the specification
$$\P = \Set(u\Cyc(\Z)),$$
which gives the bivariate exponential generating function
$$P(z,u) = \exp\bigg(u\log{1\over 1-z}\bigg) = \exp\bigg(\log{1\over (1-z)^u}\bigg)
    = \bigg({1\over 1-z}\bigg)^u\adveq$$
for the number of $n$-letter permutations with exactly $k$ cycles. We can calculate the partial derivatives
$$P_u(z,u) = \bigg({1\over 1-z}\bigg)^u\log{1\over 1-z}$$
Here the random variable $\chi$ counts the number of cycles, and we have
$$\ex_{\P_n}\{\chi\} = {[z^n] P_u(z,u)\eval_{u=1} \over [z^n]P(z,1)}
     = 1 + {1\over 2} + \cdots + {1\over n} = H_n.\adveq$$
The harmonic numbers $H_n$ can be expanded as $\log n + \gamma + O(1/n)$, so we find that there are roughly
$\log n$ cycles in a random permutation on $n$ letters.
A slightly harder analysis of Taylor coefficients retrives the second moment.
From this, it can be shown that $\var_{\P_n}\{\chi\}\sim \log n$ an well.
\medskip

\boldlabel External nodes in a binary tree. Hidden inside the specification for binary trees,
we can see that we are actually only counting internal nodes, since we have given $\E$, the external node,
the generating function of 1 (meaning it has size 0). If we mark it with a $u$,
we arrive at
$$\B = u\E + \Z\times\B\times\B\adveq$$
and the bivariate ordinary generating function satisfies
$$B(z,u) = u + zB(z)^2.$$
Using the quadratic formula once again, we find that the generating function is
$$B(z,u) = {1-\sqrt{1-4uz} \over 2z},\adveq$$
which checks out, since setting $u=1$ gives exactly \refeq{\the\catalaneq}. We have the partial derivatives
$$P_u(z,u) = {1\over \sqrt{1-4uz}} \qquad\hbox{and}\qquad P_u^2(z,u) = {2z \over (1-4uz)^{3/2}}.\adveq$$
With the substitution $u=1$,
these are both relatively well-known generating functions. The first is the generating function
for the central binomial coefficients ${2n\choose n}$ and
the second is $z {\partial\over \partial z} P_u(z,1) $, so it has coefficients $n{2n\choose n}$
(the powers of $n$ drop down into the coefficient when differentiating, and multiplying by $z$ restores
the correct power).
We calculated in \refeq{\the\catalanformula} that the
total number of binary trees with $n$ internal nodes is ${2n\choose n}/(n+1)$, so applying our method gives us
$$\ex_{\B_n}\{\chi\} = {{2n\choose n}\over {1\over n+1}{2n\choose n}} = n+1\adveq$$
and
$$\ex_{\B_n}\{\chi^2\} = {n{2n\choose n} \over {1\over n+1}{2n\choose n}} + n + 1 = n(n+1)+n+1 = (n+1)^2.$$
So the average number of external nodes in a tree with $n$ internal nodes is $n+1$ and since
$$\var_{\B_n}\{\chi\} = \ex_{\B_n}\{\chi^2\} - \ex_{\B_n}\{\chi\}^2 = (n+1)^2 - (n+1)^2 = 0,\adveq$$
we discover that, in fact, every tree with $n$ internal nodes has exactly $n+1$ external nodes.
(We have performed what is probably the most convoluted proof of this simple fact, but it is a good exercise
in calculating moments using bivariate generating functions.)
\bigskip
{\obeylines\eightssi
\hfill And if I should live to be
\hfill The last leaf upon the tree
\eightss
\smallskip
\hfill --- OLIVER WENDELL HOLMES, {\eightssi The Last Leaf} (1831)}
\bigskip

\advance\sectcount by 1
\section\llap{*}\the\sectcount. Rudiments of Complex Analysis

In this section, we will collect definitions and theorems that will be useful in studying the coefficient
asymptotics of generating functions. We will state results without proof; the details can be found any
introductory textbook on complex analysis (e.g., Flanigan (1983)). Readers who are familiar with
complex numbers may choose to skim or skip this section.

The complex plane $\CC$ is the set of all numbers $x + iy$ were $x$ and $y$ are real numbers and $i$
is the square root of $-1$. If $z = x + iy$, then the {\it real part} of $z$ is $\Re z = x$ and the
{\it imaginary part} of $z$ is $\Im z = y$. The complex conjugate of $z = x + iy$ is $x-iy$, and it is
easily verified, from the identity $i^2 = -1$, that
$$z\cdot \bar z = x^2 + y^2.$$
Letting $|z| = \sqrt{x^2 + y^2}$ denote the {\it modulus} of $z$, we find that
$${z\cdot \bar z\over |z|^2} = 1,$$
making it clear that the inverse $1/z$ of a complex number $z$ is equal to $\bar z / |z|^2$.

For real $\theta$, we have the well-known formula
$$e^{i\theta} = \cos \theta + i\sin \theta,\adveq$$
which gives another way to represent a complex number. The complex number $z = x+iy$ can also be written
as $|z|e^{i\theta}$, where $\theta$ is the angle that $z$ makes with the positive real axis.
Note that $e^{i\theta} = e^{i(\theta + 2\pi k)}$ for any integer $k$.
\bigskip
{\obeylines\eightssi
\hfill The imaginary relish is so sweet
\hfill That it enchants my sense.
\eightss
\smallskip
\hfill --- WILLIAM SHAKESPEARE, {\eightssi Twelfth Night} (1602)}
\bigskip

\boldlabel Complex differentation. A complex function defined on a domain $D$
is said to be {\it analytic} at a point $z_0\in D$ if for some open ball $\Omega$ centred at
$z_0$, $f$ is representable as a power series
$$f(z) = \sum_{n\geq 0} f_n(z-z_0)^n.$$
For example, $1/(1-z)$ is analytic for at any $z_0$ in the ball of radius 1 about the origin.
A complex function whose domain is D
is called {\it differentiable} or {\it holomorphic} at $z_0\in D$ if the limit
$$f'(z_0) = \lim_{z\to z_0} {f(z) - f(z_0) \over z- z_0} $$
exists, in the sense that $f'(z_0)$ has the same value regardless of the manner in which $z$ approaches
$z_0$ in the complex plane. The following theorem says that these two notions are equivalent.

\parenproclaim Theorem E (Basic Equivalence Theorem). A complex function $f$ defined on a domain
$D$ is analytic at a point $z_0\in D$ if and only if it is differentiable at $z_0$.\slug

It is a useful fact that if a function is differentiable at $z_0$,
then its derivative is also differentiable at $z_0$, meaning
that it admits derivatives of all orders, and we have the expansion
$$f(z) = \sum_{n\geq 0} {f^{(n)} \over n!} (z-z_0)^n.\adveq$$
Of course, if a function is analytic in a disc around $z_0$, then we can
differentiate it by taking derivatives of the terms in its power series.
\medskip

\boldlabel Complex integration. The Fundamental Theorem of Calculus states that
for a real-valued function $f$ with antiderivative $F$,
$$\int_a^b f(x)\d x = F(b) - F(a).\adveq$$
We can think of this as summing the values of $f$ over a path from $a$ to $b$. Since $\RR$ is
one-dimensional,
there is only one way to get from $a$ to $b$, but on the complex plane, we can integrate over
any path $\gamma$ from $z_0\in \CC$ to $z_1\in \CC$. We have the following complex analogue of
the Fundamental Theorem of Calculus.

\parenproclaim Theorem P (Path integration). Let $f$ a complex function,
continuous in a domain $D$, and let $F$
be a complex function, analytic in $D$, that satisfies $F'(z) = f(z)$ for all $z\in D$.
If $\gamma$ is any piecewise smooth path from $z_0\in D$ to $z_1\in D$, then
$$\int_\gamma f(z)\d z = F(z_1) - F(z_0).\noskipslug$$

We can evaluate complex integrals over paths by parametrising the curve $\gamma$ as a function
of a real variable $t$ and then performing a change of variables. If $\gamma(t)$ is a parametrised curve
and $t$ ranges from $a$ to $b$, then
$$\int_\gamma f(z)\d z = \int_a^b f\big(\gamma(t)\big)\gamma'(t)\d t.\adveq$$

As an example, if $f(z) = z^2$ and we want to find the integral of $f$ from $z_0 = 0$ to $z_1 = 1+2i$. We can
walk from 0 to $1+2i$ via a path $\gamma_1$ that goes from $0$ to $2i$ and then another path $\gamma_2$ that goes
from $2i$ to $1+2i$. So $\gamma_1(t) = it$ for $0\leq t\leq 2$ and $\gamma_2(t) = 2i + t$ for $0\leq t\leq 1$ and
we have the derivatives ${\gamma_1}'(t) = i$ and ${\gamma_2}'(t) = 1$.
Putting this all together, the integral over the
whole path is
$$\eqalign{
\int_\gamma z^2 \dz &= \int_0^2 (it)^2\cdot i\d t + \int_0^1 (2i + t)^2\d t \cr
&= \int_0^2 -it^2\d t + \int_0^1 t^2 + 4it - 4 \d t \cr
&= \bigg[-{it^3\over 3}\bigg]_{t=0}^2 + \bigg[{t^3\over 3} + 2it^2 - 4t\bigg]_{t=0}^1 \cr
&= -{8i\over 3} + {1\over 3} + 2i - 4 \cr
&= -{11\over 3} - {2i\over 3} \cr
}\adveq$$
Of course, in this example, we could have just noticed that the antiderivative
of $f(z) = z^2$ is $F(z) = z^3/3$ and applied
Theorem P to get
$$\int_\gamma f(z)\d z = F(1+2i) - F(0) = {(1+2i)^3\over 3} = -{11\over 3} - {2i\over 3}.\adveq$$

When $\gamma$ is a {\it closed Jordan curve}, i.e.\ it starts and ends at the same point and does not
cross itself, we often write the
integral sign with a circle around it.
The rules of integration are the same; the circle is just there to remind
us that our path goes around in a loop. Integrals over closed curves are important, because of the following
theorem.

\parenproclaim Theorem I (Cauchy Integral Theorem). Let $\gamma$ be a closed Jordan curve in a domain
$D$. Suppose $f$ is a function that is analytic both on $\gamma$ and the interior of $\gamma$. Then
$$\oint_\gamma f(z)\d z = 0.\noskipslug$$

A corollary of Theorem I is that if $\gamma_1$ is a closed curve that is fully contained in the interior
of another closed curve $\gamma_2$ and $f$ is analytic on both curves as well as the region between
the two curves, then
$$\oint_{\gamma_1} f(z)\d z = \oint_{\gamma_2} f(z)\d z.\adveq$$
\medskip

\boldlabel Singularities. If functions tended to be analytic everywhere on $\CC$, then Theorem I tells
us that integration wouldn't be very interesting. Luckily for us, most of the functions we study
will fail to be analytic at certain points. Let $D$ be a domain and let $z_0\in D$ be an interior point.
If $f$ is not analytic at $z_0$ but it is analytic on $D\setminus \{z_0\}$, then the point $z_0$ is said
to be a {\it singularity} of $f$. There are three types of singularities:
\medskip
\item{i)} If $\lim_{z\to z_0} f(z) = w_0$ for some finite value $w_0$, then
it is possible to continuously extend $f$ to an analytic function $g$ on all of $D$
(just set $g(z_0) = w_0$). In this case, $z_0$ is called a {\it removable singularity}.
\smallskip
\item{ii)} If $\lim_{z\to z_0} |f(z)| = \infty$, then $z_0$ is called a {\it pole} of $f$.
\smallskip
\item{iii)} If neither (i) nor (ii) hold, then $z_0$ is called an {\it essential singularity} of $f$.
\medskip
We will now show that each of these types of singularities actually pop up in the wild with some examples.

First, we can create a function with a removable singularity by taking a {\it bona fide} analytic function $f$
and consider its restriction $f'$ on the domain $D\setminus \{z_0\}$. Now $z_0$ is a removable singularity
of the function $f'$, since $f'$ can be continuously extended to an analytic function on all of $D$.

An easy example of a function with a pole is $f(z) = 1/z$. This function is analytic everywhere on $\CC$ except at
$z_0 = 0$, and $\lim_{z\to z_0} |1/z| = \infty$. This tells us that $f(z)$ may have an interesting integral
around the origin. It does not matter which closed curve we take around 0, so the unit circle $C$ is a tempting
choice. We can parametrise $C$ by $\gamma(\theta) = e^{i\theta}$ for $0\leq \theta< 2\pi$, which has the
derivative $\gamma'(\theta) = ie^{i\theta}$. We now find that
\newcount\twopii
\twopii=\eqcount
$$\oint_C {1\over z}\d z = \int_0^{2\pi} e^{-i\theta} ie^{i\theta} \d\theta= i\int_0^{2\pi} 1\d\theta = 2\pi i.\adveq$$
This integral crops up a lot in complex analysis; commit it to memory!

A function with an essential singularity is the square root function $\sqrt z$. This is because, given
$z = re^{i\theta}$, we can rewrite
$z = re^{i\theta + 2\pi ik}$ for any integer $k$, so for any integer $k$, $w = \sqrt re^{i\theta/2 + \pi i k}$
satisfies the property that $w^2 = z$. To get a well-defined square root function, we set
$$\sqrt z = \sqrt re^{i\theta/2},\adveq$$
where $\theta\in (-\pi, \pi]$. This spells big trouble for complex numbers on the negative real axis.
For example, let $z = -1 = e^{i\pi}$. Note that $\sqrt{e^{i(\pi - \eps)}} = e^{i(\pi/2 - \eps/2)}$ and
$\sqrt{e^{-i(\pi -\eps)}} = e^{-i(\pi/2 - \eps/2)}$ (so both angles are in the range $(-\pi, \pi]$.
As $\eps\to 0$, the arguments to the function both
get arbitrarily close to $-1$ but
the outputs have different limits ($i$ and $-i$). So the square root function has an infinite
number of essential singularities (any negative real number); these are often called {\it algebraic
singularities}.

Another function with essential singularities is the natural logarithm $\log z$. Note that, given
$z = re^{i\theta}$, the number $w = \log r + 2k\pi i\theta$ satisfies $e^w = z$ for any integer $k$.
Once again, to get a well-defined function we must restrict $\theta$ to a single loop around the unit circle:
$$\log z = \log r + i\theta,\adveq$$
where $\theta\in (-\pi, \pi]$. Now, just as with the square root, we have a discontinuity in the argument
for all real numbers $>1$. These are the {\it logarithmic singularities}.
\medskip

\boldlabel Residues. If we allow a power series expansion around a point $z_0$
to have a finite number of negative powers, then it is
called a {\it Laurent series}; this has the form
$$f(z) = \sum_{n\geq -M} f_n (z-z_0)^n = {f_{-M} \over (z-z_0)^M} + \cdots + {f_{-1}\over (z-z_0)} + f_0
    + f_1(z-z_0) + f_2(z-z_0)^2 + \cdots,$$
for some nonnegative integer $M$. Note that, integrating on a circle of radius 1 around $z_0$, we
can make the substitution $z - z_0 = e^{i\theta}$ to get $dz = ie^{i\theta}\d\theta$ and
$$\oint (z-z_0)^M \d z = i\int_0^{2\pi} e^{i(M+1)\theta}\d\theta
    = \cases{2\pi i & if $M=-1$; \cr 0 & otherwise.}\adveq$$
So the coefficient $f_{-1}$ of $(z-z_0)^{-1}$ is special; it is the only term that survives the integration
of a Laurent series. We call this the {\it residue of $f$ at $z_0$}
and denote it by $\Res(f,z_0)$; we have
$$\Res(f,z_0) = {1\over 2\pi i} \oint f(z)\d z.\adveq$$
This is the crucial fact that will allows the following theorem to extract coefficients of generating functions.

\parenproclaim Theorem C (Cauchy coefficient formula). Let $f(z)$ be analytic. The coefficient
of $z^n$ is given by the formula
$$[z^n]f(z) = {1\over 2\pi i} \oint {f(z) \over z^{n+1}}\d z,\adveq$$
where the integral is taken on any closed curve about the origin.\slug

\advance\sectcount by 1
\section\the\sectcount. Generating Functions as Analytic Functions

This section represents a turning point in our study of generating functions. Up until now, we have treated
generating functions as purely formal objects. However, we can get a lot of information about the asymptotic
behaviour of a power series' coefficients by treating it as a function of a complex argument. This section
will give a general feel for how to do so, without getting too formal. A key theorem that allows us to work
with generating functions as analytic objects is the following.

\parenproclaim Theorem P (Pringsheim). If $f(z)$ has a series expansion whose coefficients are all
nonnegative and the radius of convergence of $f(z)$ is $R$, then the real number $R$ is a singularity
of $f(z)$.\slug

Since counting objects usually induces a generating function with nonnegative coefficients, this theorem
applies to most of the interesting cases we study.
Note that we will only deal with the case where there is a {\it unique}
root of smallest modulus, meaning that $R$ is the only singularity on the circle $|z| = R$. This assumption
eliminates complications that involve periodicities in the coefficients.
\medskip

\boldlabel Rational functions. Treating a generating function as a complex function pays off immediately
when the function is a ratio $f(z)/g(z)$ of two polynomials. Since a polynomial of degree $n$ always has
$n$ complex roots, we can factorise $g(z)$ into a product of linear factors and then use the partial fraction
expansion to extract the coefficients. For example, if $f(z) = z$ and $g(z) = 18z^3 - 3z^2 - 4z + 1$, then
$$\eqalign{
{f(z)\over g(z)} &= {z\over 18z^3 - 3z^2 - 4z + 1} = {z\over 18(z-1/3)^2(z+1/2)} \cr
&= {1\over 45(z-1/3)^2} + {1\over 25(z-1/3)} - {1\over 25(z+1/2)}\cr
&= {1\over 5(1-3z)^2} - {3\over 25(1-3z)} - {2\over 25(1+2z)}.\cr
}\adveq$$
This immediately tells us that
\newcount\exactrat
\exactrat=\eqcount
$$\eqalign{
[z^n]{f(z)\over g(z)} &= {1\over 5}(n+1)3^{n} - {3\over 25} 3^n - {2\over 25}(-2)^n \cr
&=  {1\over 5}n3^n + {2\over 25} 3^n - {2\over 25}(-2)^n. \cr
}\adveq$$
Notice that only the first term is important, because $3^n$ is exponentially larger than $2^n$. We got
an exponential factor of $\sim 3^n$ because the root of $g(z)$ that is closest to the origin is $1/3$. The fact
that this root has multiplicity $2$ gives a factor of $\sim n$ in the first term. It is safe to ignore
all the other terms because they are much smaller, as formalised in the following theorem.

\parenproclaim Theorem G (Rational asymptotics). Let $f(z)$ and $g(z)$ be polynomials that are relatively prime,
with $g(z) \neq 0$. If $g(z)$ has a a unique real root $\alpha$ of smallest modulus whose multiplicity is $r$, then
$$[z^n]{f(z)\over g(z)} \sim C\alpha^{-n} n^{r-1} \qquad\hbox{where}\qquad
   C = {r(-1/\alpha)^r f(\alpha) \over g^{(r)}(\alpha)}.\noskipslug$$

This will become clear later on when we discuss a more general result.
Applying this theorem to our most recent example, we have $\alpha = 1/3$, $r = 2$, and
$g''(z) = 108z-6$. So we can derive
$$[z^n]{z\over 18z^3 - 3z^2 - 4z + 1} = {2(-3)^2 (1/3) \over 108(1/3)-6} 3^n n^{2-1} = {1\over 5} n3^n,\adveq$$
which is exactly the first term in \refeq{\the\exactrat}.
\medskip

\boldlabel Fibonacci asymptotics. As an example, recall the generating
function for the Fibonacci numbers, $z/(1-z-z^2)$. This function has poles at
$$-\phi = {-1-\sqrt 5\over 2}\approx -1.618 \qquad\hbox{and}\qquad
   \bar\phi = {-1+\sqrt 5\over 2}\approx0.618,\adveq$$
where $\phi$ is the golden ratio.
The root of smaller modulus is $\alpha = \bar\phi = 1/\phi$, and the multiplicity $r = 1$. This gives
$$F_n = [z^n]{z\over 1-z-z^2} \sim {(-\phi)(1/\phi)\over 1-2/\phi} \phi^n
    = {\phi \over \phi + 2}\phi^n = {1 + \sqrt 5 \over 5 + \sqrt 5}\phi^n = {\phi ^n\over \sqrt 5},\adveq$$
which says that the ratio between successive Fibonacci numbers tends towards $\phi$.
\medskip

\boldlabel Meromorphic functions. Rational functions are special cases of a broader class of functions
called meromorphic functions. A complex function is {\it meromorphic} if it can be expressed as the ratio
of two analytic functions. A function is meromorphic in a region if and only if it is analytic except for
at a finite set of poles.
We will use the fact that any function that is meromorphic at the point $z_0$
admits a Laurent expansion
$$h(z) = {h_{-M} \over (z-z_0)^M} + \cdots + {h_{-1}\over (z-z_0)} + h_0
    + h_1(z-z_0) + h_2(z-z_0)^2 + \cdots,$$
where $M$ is a nonnegative integer. We say that $h$ has a {\it pole of order $M$} at the point $z_0$.
In this case, the function $(z-z_0)^M h(z)$ is analytic at $z_0$, because the Laurent series becomes
a common-or-garden Taylor series. We have the following generalisation of Cauchy's coefficient formula
for meromorphic functions.

\parenproclaim Theorem R (Residue theorem). Let $h(z)$ is meromorphic in a region $D$ and let
$\gamma$ be a closed loop in $D$. Suppose that $\gamma$ encloses $k$ poles of $h$; call them
$\alpha_1,\ldots,\alpha_k$. Then
$$[z^n]h(z) = {1\over 2\pi i} \oint_\gamma h(z)\d z = \sum_{i=1}^k \Res(h, \alpha_i).\adveq$$
In particular, if $h$ is meromorphic in closed ball $|z|\leq R$, analytic at 0,
and the poles have orders $m_1,\ldots,m_k$, then
$$[z^n]h(z) = {p_1(n)\over {\alpha_1}^n} + \cdots + {p_k(n)\over {\alpha_k}^n} + O(R^{-N}),\adveq$$
where each polynomial $p_i$ has degree $m_i - 1$.\slug

As before, we can get a very good approximation by looking only at the pole $\alpha$ of smallest modulus.
The residue around that pole contributes exponentially more to the coefficient asymptotics than the others,
so we should analyse the Laurent expansion of $h(z)$ around $\alpha$. If $\alpha$ has order $M$, then
this looks like
$$h(z) = {h_{-M} \over (z-\alpha)^M} + \cdots + {h_{-1}\over (z-\alpha)} + h_0
    + h_1(z-\alpha) + h_2(z-\alpha)^2 + \cdots,$$
and when $z$ is very close to $\alpha$, only the first term matters. In other words,
$$h(z) \sim {h_{-M} \over (z-\alpha)^M} = {h_{-M}\over (-\alpha)^M (1-z/\alpha)^M}
   = (-1)^M{h_{-M}\over\alpha^M} \sum_{n\geq 0} {n+M-1\over \alpha^n}z^n.\adveq$$
The actual residue $h_{-M}$ can be computed as a limit, by repeated application of l'Hospital's Rule.
If $h(z)$ is equal to the ratio $f(z)/g(z)$ of analytic functions, then
$$h_{-M} = \lim_{z\to \alpha} {(z-\alpha)^M f(z) \over g(z)} = {Mf(\alpha) \over g^{(M)}(\alpha)}.\adveq$$
Putting this all together, we have the following theorem.

\parenproclaim Theorem M (Meromorphic asymptotics). If $h(z) = f(z)/g(z)$ is a meromorphic function with
a unique pole of smallest modulus $\alpha$ whose order is $M$, then
$$[z^n]h(z) \sim (-1)^M {Mf(\alpha) \over \alpha^M g^{(M)}(\alpha)} \alpha^{-n} n^{M-1}.\noskipslug$$

We now see that Theorem G is a special case of this, since a multiplicity-$r$
root of a polynomial in the denominator induces a pole of order $r$.
\medskip
